---
title: "Data Cleaning and Standards Assignment"
author: "Katie Danyk"
date: "2026-09-28"
output: html_document
---

# Step 1: Load the required packages and read in the raw data

Before I begin the assignment, I need to load the packages and ensure that my code is reproducible by recording package versions and establishing file paths.


### 1a. Load packages

```{r, message = FALSE}
library(renv) 
library(here)
library(tidyverse)
library(assertr)
library(skimr)
library(stringdist)
```


### 1b. Record and lock package versions

I will do this with the 'init()', 'snapshot()', and 'status()' commands from the package 'renv'
```{r}
#renv::init()
snapshot()
status()

#renv::restore()
```


### 1c. Build file paths

I will do this using the 'here' package
```{r}
here::here(
  "Raw_data",
  "chl-a-samples-messy.csv"
)

here::here(
  "Raw_data",
  "stations-messy.csv"
)

here::here(
  "Raw_data",
  "waterbodies-messy.csv"
)
```


### 1d. Read in raw data

I will use the 'read_csv' command from readr
```{r}
# messy chl a samples data
chl_a_messy <- read_csv("Raw_data/chl-a-samples-messy.csv")
  
# messy stations data
stations_messy <- read_csv("Raw_data/stations-messy.csv")
  
# messy waterbodies data 
waterbodies_messy <- read_csv("Raw_data/waterbodies-messy.csv")
```




# Step 2: Explore the datasets

Before I can clean the data, I need to explore the structure, variable classes, distributions, and categorical values of each dataset. I also need to use an entity relationship diagram (ERD) to inspect the column names in each of the datasets and determine which shared identifier columns can be used to connect the datasets. 


### 2a. Chl a dataset

First, I'm going to look at the structure, variable classes, and categorical values of the chl a dataset. I'm going to use 'str()' to see the class of each variable and 'dim()' to get the number of rows and columns. I'm also going to use 'head()' to look at the first five rows of data to get an idea of what they look like and 'skim()' to generate a summary of the dataset and each variable.
```{r}
str(chl_a_messy)

dim(chl_a_messy)

head(chl_a_messy)

skim(chl_a_messy)
```

~~ SUMMARY~~ 
*str() output*
From the str() output, I can see that the variables with the class 'character' are: 
  - chl_a_sample_code
  - station_code
  - date
  - sample_volume_filtered_ml
  - absorbance_663nm
  - description
  - unit

I can also see that the variables with the class 'numeric' are: 
  - sample_replicate
  - subsample_replicate
  - extract_volume_ml
  - absorbance_750nm
  - absorbance_645nm
  - absorbance_630nm
  - chl_a_16ed
  - chl_a_20ed

Based on the Data Dictionary, and the output of the str() command, I can tell that some of the variables are classed incorrectly:
  - sample_volume_filtered_ml should be 'numeric'
  - absorbance_663nm should be class 'numeric'
  - sample_replicate should be class 'integer' because it can only be whole numbers
  - subsample_replicate should be class 'integer' because it can only be whole numbers
  - date should be class 'date'
  
*dim() output*
The dim() output tells me that there are 1578 rows and 15 columns.

*skim() output*
From the skim() output, I can see that none of the 'character' variables have white spaces or missing values, except "description", but this is okay because the Data Dictionary says that "description" was not always recorded. There are also two unique values for "units", when ther should only be one.

For the 'numeric' variable part of the skim() output, I can see that the only variables with missing values are "sample_replicate" and "subsample_replicate". This is not a problem for "subsample_replicate" based on the Data Dictionary description, but the Data Dictionary does not mention missing values for "sample_replicate" so there are potentially errors here.

I can also see that "extract_volume_ml" has a -9.8 in the "p0" column, leading me to believe there is a typo in the "extract_volume_ml" column because this column should only contain positive values based on the Data Dictionary description.

Lastly, the "p100" column has a 70.680 and 391.050 for "chl_a_16ed" and "chl_a_20ed" respectively, which both seem a bit high compared to the means and sd's and suggests that there might be typos in these columns.



Then, I'm going to look at the distributions of the 'numeric' variables with histograms 
```{r}
ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = extract_volume_ml))

ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = as.numeric(sample_volume_filtered_ml))) # I'm coercing the variable to 'numeric' inside the plot only, just so I can see the distribution

ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = absorbance_750nm))

ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = as.numeric(absorbance_663nm))) # I'm coercing the variable to 'numeric' inside the plot only, just so I can see the distribution

ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = absorbance_645nm))

ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = absorbance_630nm))

ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = chl_a_16ed))

ggplot() +
  geom_histogram(data = chl_a_messy, aes(x = chl_a_20ed))
```

~~ SUMMARY ~~
*extract_volume_ml*
There is a negative value, which I also identified with the skim() command above. This must be a typo as these values should all be positive based on the Data Dictionary. 

*sample_volume_filtered_ml*
There is a value around 10,000 which is likely a typo because a value this large doesn't make sense based on the Data Dictionary. I also had to coerce this variable to numeric, resulting in a warning message. This tells me there is likely a comma or a letter accidentally included in one of the entries that I will have to correct.

*absorbance_750nm*
This distribution looks alright, although there is potentially an outlier around 0.20. I will formally check for outliers in a later step.

*absorbance_663nm*
This distribution looks alright. I also had to coerce this variable to numeric, resulting in a warning message. This tells me there is likely a comma or a letter accidentally included in one of the entries that I will have to correct.

*absorbance_645nm*
This distribution looks alright, but I will formally check for outliers in a later step.

*absorbance_630nm*
This distribution looks alright, but I will formally check for outliers in a later step.

*chl_a_16ed*
This distribution looks alright, although there could be outliers around 65 and 70. I will formally check for outliers in a later step.

*chl_a_20ed*
There is a value around 400, which is potentially a typo and likely an outlier. I will formally check for outliers in a later step.



### 2b. Stations dataset

First, I'm going to look at the structure, variable classes, and categorical values of the chl a dataset. I'm going to use 'str()' to see the class of each variable and 'dim()' to get the number of rows and columns. I'm also going to use 'head()' to look at the first five rows of data to get an idea of what they look like and 'skim()' to generate a summary of the dataset and each variable.
```{r}
str(stations_messy)

dim(stations_messy)

head(stations_messy)

skim(stations_messy)
```

~~ SUMMARY~~ 
*str() output*
From the str() output, I can see that the variables with class 'character' are:
  - station_code
  - waterbody_code
  - station_name
  - station_type
  
I can also see that the variables with class 'numeric' are: 
  - latitude
  - longitude
  
All the classes are correct

*dim() output*
From the dim() output, I can see that there are 97 rows and 6 columns

*skim() output*
For the 'character' part of the skim() output, I can see that none of the columns contain missing values or white spaces, which is correct based on the Data Dictionary descriptions.

In the 'numeric' part of the output, I can see that there are no missing values, which is correct based on the Data Dictionary descriptions. 

There seems to be typos in both the "latitude" and "longitude" columns. In the skim() output, "p0" for "latitude" is -83.96436, which does not make sense for where Ontario is in the world. Additionally, "p100" for "longitude" is 84.11925, which also does not make sense for where Ontario is in the world.



Then, I'm going to look at the distributions of the 'numeric' variables with histograms
```{r}
ggplot() +
  geom_histogram(data = stations_messy, aes(x = latitude))

ggplot() +
  geom_histogram(data = stations_messy, aes(x = longitude))
```

~~ SUMMARY ~~
*latitude*
I can see the value at about -84 again that I previously identified using skim()

*longitude*
It looks like there are two typos in "longitude". I previously identified the one at about 84 with skim(), but I can see an additional point here at about 50. The point at about 80 is likely just missing the "-" and can be fixed later. It's possible the latitude typo at -84 and the longitude typo at about 50 are coordinates of one station, just entered in the incorrect column, but I will check this in a later step.



### 2c. Waterbodies dataset

First, I'm going to look at the structure, variable classes, and categorical values of the chl a dataset. I'm going to use 'str()' to see the class of each variable and 'dim()' to get the number of rows and columns. I'm also going to use 'head()' to look at the first five rows of data to get an idea of what they look like and 'skim()' to generate a summary of the dataset and each variable.
```{r}
str(waterbodies_messy)

dim(waterbodies_messy)

head(waterbodies_messy)

skim(waterbodies_messy)
```

~~ SUMMARY~~ 
*str() output*
From the str() output, I can see that the columns with class 'character' are:
  - waterbody_code
  - waterbody_name
  
Both columns have the correct class based on the Data Dictionary descriptions

*dim() output*
The dim() output lets me knwo that this dataset has 87 rows and 2 columns

*skim() output*
The skim() output tells me that there are no missing values or white spaces in either column, which is correct based on the Data Dictionary.

I can also see that "n_unique" is 87 for both "waterbody_code" and "waterbody_name". I know this is correct because each waterbody should have a unique and code



I cannot look at the distribution of these variables because they are not numeric.



### 2d. ERD

From the provided ERD, I can see that the column "waterbody_code" can be used to join the waterbodies and stations datasets. From my skim() output above, I can see that the "waterbody_code" contains 87 unique values in both datasets.

Once the waterbodies and stations datasets are merged, I can then use the column "station_code" can be used to join the previously merged datasets with the chl a dataset. From my skim() output about, I can see that the stations dataset has 97 unique "station_code" values, while the chl a dataset only has 70 unique values. This is likely because not every station was sampled, but I will check that there are no incorrect station codes in a later step.





# Step 3: Cleaning

### 3a. Chl a dataset

From my exploration of the chl a dataset above, I have identified the following errors:
  *Individual outliers*
    - potential outlier in absorbance_750nm around 0.20
    - potential outliers in chl_a_16ed around 65 and 70
    - potential outlier in chl_a_20ed around 400
  
  *Formatting errors*
    - sample_volume_filtered_ml should be 'numeric'
    - absorbance_663nm should be class 'numeric'
    - sample_replicate should be class 'integer' because it can only be whole numbers
    - subsample_replicate should be class 'integer' because it can only be whole numbers
    - date should be class 'date' and changed to ISO 8601
    - chl_a_16ed and chl_a_20ed contain two different units when they should only be mg/L
    
  *String errors*  
    - no previously identified errors
    
  *Invalid values*
    - extract_volume_ml has a negative value
    - sample_volume_filtered_ml has a value around 10,000
  
More errors may be revealed as I clean the data and they will be dealt with as they appear. I'm going to start with resolving *formatting errors* because I believe these will be the easiest to resolve.

Before I begin, I will duplicate the dataset so I'm not editing the raw data
```{r}
chl_a_clean <- chl_a_messy
```

First, sample_volume_filtered_ml should be 'numeric' not 'character'. This leads me to believe that there are embedded units or a comma instead of a decimal. I'm going to check for these using the 'grep()' command from base R.
```{r}
# step 1. identify the row that contains the typo
grep(',|mL|ml', chl_a_messy$sample_volume_filtered_ml) # identifies a match in row 163

# step 2. view the row that contains the typo so I can avoid hard coding the fix
chl_a_messy %>%
  slice(163) 

# step 3. replace the incorrect value with the correct value
chl_a_clean$sample_volume_filtered_ml[chl_a_clean$sample_volume_filtered_ml == '1,000 mL'] <- '1000'
 
# step 4. check to make sure it worked 
grep(',|mL|ml', chl_a_clean$sample_volume_filtered_ml) # identifies no matches

# step 5. change the class to 'numeric'
chl_a_clean$sample_volume_filtered_ml <- as.numeric(chl_a_clean$sample_volume_filtered_ml)

# step 6. check that the change worked
is.numeric(chl_a_clean$sample_volume_filtered_ml)
```

Next, absorbance_663nm should be 'numeric' not 'character. This leads me to believe that there are embedded units or a comma instead of a decimal. I'm going to check for these using the 'grep()' command from base R.
```{r}
# step 1. identify the row that contains the typo
grep(',|nm', chl_a_messy$absorbance_663nm) # identifies a match in row 154

# step 2. view the row that contains the typo so I can avoid hard coding the fix
chl_a_messy %>%
  slice(154) 

# step 3. replace the incorrect value with the correct value
chl_a_clean$absorbance_663nm[chl_a_clean$absorbance_663nm == '0,026'] <- '0.026'

# step 4. check to make sure it worked 
grep(',|nm', chl_a_clean$absorbance_663nm) # identifies no matches

# step 5. change the class to 'numeric'
chl_a_clean$absorbance_663nm <- as.numeric(chl_a_clean$absorbance_663nm)

# step 6. check that the class is correct
is.numeric(chl_a_clean$absorbance_663nm)
```

Next, sample_replicate should be class 'integer' not 'numeric'. I will look for decimals using the 'grep()' command from base R.
```{r}
# step 1. identify the row that contains the typo
grep('.', chl_a_messy$sample_replicate) # there doesn't appear to be a decimal present so I should be okay to just switch the class

# step 2. change the class to 'integer'
chl_a_clean$sample_replicate <- as.integer(chl_a_clean$sample_replicate)

# step 3. check that the class is correct
is.integer(chl_a_clean$sample_replicate)
```

Next, subsample_replicate should be class 'integer' not 'numeric'. I will look for decimals using the 'grep()' command from base R. 
```{r}
# step 1. identify the row that contains the typo
grep('.', chl_a_messy$subsample_replicate) # there doesn't appear to be a decimal present so I should be okay to just switch the class

# step 2. change the class to 'integer'
chl_a_clean$subsample_replicate <- as.integer(chl_a_clean$subsample_replicate)

# step 3. check that the class is correct
is.integer(chl_a_clean$subsample_replicate)
```

Next I will address the date. Before I can change the class to 'date', I need to make sure that all the dates are formatted in the same way. Incorrectly formatted dates could include "/", " ", "_", ",", or words
```{r}
# step 1. determine if there are any incorrectly formatted dates
grep('/| |_|,', chl_a_messy$date) # identified three incorrect dates in rows 507, 524, and 1008

# step 2. view the rows that contain the typos so I can avoid hard coding the fix
chl_a_messy %>%
  slice(507, 524, 1008) 

# step 3. replace the incorrectly formatted dates with yyyy-mm-dd format
chl_a_clean$date[chl_a_clean$date == 'June 14, 1999'] <- '1999-06-14'
chl_a_clean$date[chl_a_clean$date == '28/06/1999'] <- '1999-06-28'
chl_a_clean$date[chl_a_clean$date == '10/15/2002'] <- '2002-10-15'

# step 4. check to make sure it worked 
grep('/| |_|,', chl_a_clean$date) # identifies no matches

# step 5. change the class to 'date'
chl_a_clean <- chl_a_clean %>%
  mutate(date = lubridate::ymd(date))

# step 6. check that the class is correct
is.Date(chl_a_clean$date)
```

The last formatting error I identified was that two units were being used for chl_a_16ed and chl_a_20ed when there should only be one. To fix this I will convert all the mg/m3 values to mg/L. 
1 mg/m3 = 0.001 mg/L so I need to multiply these values by 0.001 to convert to m/L
```{r}
# step 1. separate the dataset into two by on the unit that was used
chl_a_messy_mgL <- chl_a_messy %>%
  filter(unit == "mg/L")
chl_a_messy_mgm3 <- chl_a_messy %>%
  filter(unit == "mg/m3")

# step 2. correct the values in the mg/m3 dataset
chl_a_messy_mgm3 <- chl_a_messy_mgm3 %>%
  mutate(chl_a_16ed_mgL = chl_a_16ed*0.001, # calculate correct values
         chl_a_20ed_mgL = chl_a_20ed*0.001,
         unit = "mg/L") %>% # change to the correct units
  dplyr::select(!c(chl_a_16ed, chl_a_20ed)) %>% # remove the columns with the incorrect units
  rename(chl_a_16ed = chl_a_16ed_mgL, # rename the correct columns so they match the original dataset
         chl_a_20ed = chl_a_20ed_mgL)

# step 3. bind the two datasets back together
chl_a_messy_mgL_v2 <- bind_rows(chl_a_messy_mgL, chl_a_messy_mgm3) # the number of observations in this new dataset is 1578, which matches the original dataset

# step 4. extract the two chl_a columns and the unit column
chl_a_cols <- chl_a_messy_mgL_v2 %>%
  select(chl_a_16ed, chl_a_20ed, unit) %>%
  rename(chl_a_16ed_new = chl_a_16ed, # rename the columns so I can bind them without the names changing automatically
         chl_a_20ed_new = chl_a_20ed,
         unit_new = unit) 

# step 5. bind these columns onto the clean dataset and remove the existing columns
chl_a_clean_v2 <- bind_cols(chl_a_clean, chl_a_cols) %>%
  dplyr::select(!c(chl_a_16ed, chl_a_20ed, unit)) %>% # remove the old columns with the incorrect units included
  rename(chl_a_16ed = chl_a_16ed_new, # rename back to the original columns names for consistence
         chl_a_20ed = chl_a_20ed_new, 
         unit = unit_new)

# step 6. check that the new clean dataset has the same dimensions as the messy dataset
chl_a_clean_v2 %>% 
  assertr::verify(nrow(.) == 1578) # no error message means there are the same number of rows
chl_a_clean_v2 %>% 
  assertr::verify(ncol(.) == 15) # no error message means there are the same number of columns
```



Now, that I've resolved all the formatting errors I originally identified, I am going to move on to the *invalid values*.

First, I will deal with the negative value is in extract_volume_ml. To make sure there are no other negatives than the one I previously identified, I will use the 'assert(within_bounds())' function from the assertr package.
```{r}
# step 1. look for negative values in all numeric/integer columns
chl_a_clean_v2 %>% 
  assert(within_bounds(0, Inf), 
         -c(chl_a_sample_code, station_code, date, sample_replicate, subsample_replicate, description, unit)) # only identified one negative value (-9.8)

# step 2. correct the negative value
chl_a_clean_v2$extract_volume_ml[chl_a_clean_v2$extract_volume_ml == '-9.8'] <- 9.8

# step 3. check that the negative value was corrected
chl_a_clean_v2 %>% 
  assert(within_bounds(0, Inf), 
         -c(chl_a_sample_code, station_code, date, sample_replicate, subsample_replicate, description, unit)) # no error message means that the negative value was corrected

# step 4. check that the replacement did not change the class
is.numeric(chl_a_clean_v2$extract_volume_ml)
```
In this situation, I chose to replace the negative value with the positive equivalent, because it seemed likely that it was a typo.

Next, I will address the value around 10,000 in sample_volume_filtered_ml. Based on the Data Dictionary, the sample_volume_filtered_ml should be between 0 and 1000, so I will check for any values > 1000 using the 'assert(within_bounds())' function from the assertr package. 
```{r}
# step 1. look for negative values in all numeric/integer columns
chl_a_clean_v2 %>% 
  assert(within_bounds(0, 1000), sample_volume_filtered_ml) # only identified the value I had found before

# step 2. correct the value that's too large
chl_a_clean_v2$sample_volume_filtered_ml[chl_a_clean_v2$sample_volume_filtered_ml == '10000'] <- 1000

# step 3. check that the negative value was corrected
chl_a_clean_v2 %>% 
  assert(within_bounds(0, 1000), sample_volume_filtered_ml) # no error message means that the value was corrected

# step 4. check that the replacement did not change the class
is.numeric(chl_a_clean_v2$sample_volume_filtered_ml)
```
In this situation, I chose to replace the 10000 with 1000, because it seemed likely that it was a typo.



Now that I've dealt with all the formatting issues and invalid values, I am able to check for *outliers*. 

First, I am going to check for outliers in absorbance_750nm using the 'insist(within_n_mads(3))' command from assertr. I am going to use 3 MAD from the median as my identifier for outliers.
```{r}
# step 1. identify potential outliers
chl_a_clean_v2 %>% 
  insist(within_n_mads(3), absorbance_750nm) # 161 potential outliers

# step 2. plot the distribution again
ggplot(chl_a_clean_v2, aes(x = absorbance_750nm)) +
  geom_histogram()
```
I am not going to remove any of the potential outliers identified by insist(within_n_mads(3)). I chose to retain them all because there are quite a few points in the upper tail of the distribution, which leads me to believe that these are unusual, but correct, values.

Next, I am going to look for outliers in chl_a_16ed using the 'insist(within_n_mads(3))' command from assertr. I am going to use 3 MAD from the median as my identifier for outliers.
```{r}
# step 1. identify potential outliers
chl_a_clean_v2 %>% 
  insist(within_n_mads(3), chl_a_16ed) # 116 potential outliers

# step 2. plot the distribution again
ggplot(chl_a_clean_v2, aes(x = chl_a_16ed)) +
  geom_histogram()
```
I am not going to remove any of the potential outliers identified by insist(within_n_mads(3)). I chose to retain them all because there are quite a few points in the upper tail of the distribution, which leads me to believe that these are unusual, but correct, values. 

Next, I am going to look for outliers in chl_a_20ed using the 'insist(within_n_mads(3))' command from assertr. I am going to use 3 MAD from the median as my identifier for outliers.
```{r}
# step 1. identify potential outliers
chl_a_clean_v2 %>% 
  insist(within_n_mads(3), chl_a_20ed) # 117 potential outliers

# step 2. plot the distribution again
ggplot(chl_a_clean_v2, aes(x = chl_a_20ed)) +
  geom_histogram()

# step 3. identify the outlier
chl_a_clean_v2 %>%
  arrange(desc(chl_a_20ed)) %>%
  slice_head(n = 5) # 391.05

# step 4. remove the outlier
chl_a_clean_v2 <- chl_a_clean_v2 %>%
  filter(!chl_a_20ed == "391.05")
```
I chose to remove one of the potential outliers identified by insist(within_n_mads(3)). I chose to remove the value at 391.05 because there was a gap of over 300 units between the highest and second highest value. 



Finally, I will check for *string errors* in the 'character' columns
```{r}
# step 1a. check if all station codes start with the correct string
as_tibble(str_detect(chl_a_messy$station_code, '\\bST0')) %>% 
  mutate(ID = row_number()) %>%
  filter(!value == "TRUE") # there is one station code that start with "ST0" in row 271

# step 1b. view the row with the incorrect station code
chl_a_messy %>%
  slice(271) # the station code is ST999 - this is likely a typo and should be corrected to ST099

# step 1c. correct the type found in step 1a
chl_a_clean_v2$station_code[chl_a_clean_v2$station_code == 'ST999'] <- 'ST099'

# step 1d. check that the typo is now correct
grep('ST999', chl_a_clean_v2$station_code) # the typo is no longer present

# step 2. check that all sample codes start with the correct string
as_tibble(str_detect(chl_a_messy$chl_a_sample_code, '\\bCH')) %>% 
  filter(!value == "TRUE") # all sample codes start with the correct string

# step 3. look for spaces in columns where there shouldn't be 
as_tibble(str_detect(chl_a_messy$station_code, '[[:space:]]')) %>% 
  filter(!value == "FALSE") # no spaces
as_tibble(str_detect(chl_a_messy$chl_a_sample_code, '[[:space:]]')) %>% 
  filter(!value == "FALSE") # no spaces

# step 4. look for lowercase letters in columns where there shouldn't be 
as_tibble(str_detect(chl_a_messy$station_code, '[[:lower:]]')) %>% 
  filter(!value == "FALSE") # no lowercases
as_tibble(str_detect(chl_a_messy$chl_a_sample_code, '[[:lower:]]')) %>% 
  filter(!value == "FALSE") # no lowercases
```


### 3b. Stations dataset

From my exploration of the stations dataset above, I have identified the following errors:
  *Individual outliers*
    - no previously identified errors 
  
  *Formatting errors*
    - no previously identified errors 
    
  *String errors*  
    - no previously identified errors
  
  *Invalid values*
    - latitude of -83.96436
    - longitude of 84.11925
    - longitude around 50

More errors may be revealed as I clean the data and they will be dealt with as they appear. I'm going to start with resolving the *invalid values*.

Before I begin, I will duplicate the dataset so I'm not editing the raw data
```{r}
stations_clean <- stations_messy
```

Because the latitudes and longitudes have an unknown number of decimal places, I am not able to replace values the way I was doing it above. Instead, I will first determine all invalid values and then separate them into a new dataset so I can deal with them appropriately
```{r}
# step 1. identify all invlalid values
stations_messy %>% 
  assert(within_bounds(0, Inf), latitude) # invalid value at row 52: -83.96436
stations_messy %>% 
  assert(within_bounds(-Inf, 0), longitude) # invalid value at row 5: 84.11925; row 52: 46.73345

# step 2. determine if longitude = 84.11925 is a typo
stations_messy %>%
  group_by(waterbody_code) %>%
  summarise(mean = mean(longitude)) %>%
  arrange(desc(mean)) # a station at -84.11925 would be very close to waterbodies, therefore it is likely a typo

# step 3. separate invalid values into new datasets
st_5_lon <- stations_messy %>%
  filter(station_code == "ST005") # the longitude will have to me multiplied by -1
st_52_lat_lon <- stations_messy %>%
  filter(station_code == "ST052") # the longitude and latitude will have to be renamed to the other name

# step 4. fix invlaide values
st_5_lon <- st_5_lon %>%
  mutate(longitude = longitude * -1)
st_52_lat_lon <- st_52_lat_lon %>%
  rename(longitude = latitude,
         latitude = longitude)

# step 5. remove the stations with invalid lat and lon valued from the clean dataset
stations_clean <- stations_clean %>%
  filter(!station_code %in% c("ST005", "ST052"))

# step 6. bind the fixed rows to the clean dataset
stations_clean_v2 <- bind_rows(stations_clean, st_5_lon, st_52_lat_lon)

# step 7. check that the dimensions of the new clean dataset are the same as the messy dataset
stations_clean_v2 %>% 
  verify(nrow(.) == 97) # no error message means the number of rows is the same
stations_clean_v2 %>% 
  verify(ncol(.) == 6) # no error message means the number of columns is the same

# step 8. check that there are no invalid lat and lon values in the clean dataset
stations_clean_v2 %>% 
  assert(within_bounds(0, Inf), latitude) # no error means no invalid values
stations_clean_v2 %>% 
  assert(within_bounds(-Inf, 0), longitude) # no error means no invalid values
```



Now that I have dealt with the issues I was able to find during exploration, I also want to check if there are any *string errors* in the 'character' columns.
```{r}
# step 1. look for spaces in columns where there shouldn't be 
str_detect(stations_messy$station_code, '[[:space:]]') # no spaces
str_detect(stations_messy$waterbody_code, '[[:space:]]') # no spaces
str_detect(stations_messy$station_type, '[[:space:]]') # no spaces

# step 2. look for lowercase letters in columns where there shouldn't be 
str_detect(stations_messy$station_code, '[[:lower:]]') # no lowercases
str_detect(stations_messy$waterbody_code, '[[:lower:]]') # no lowercases

# step 3a. use a cluster diagram to look for typos in station_name
dist.matrix.name <- stringdist::stringdistmatrix(stations_messy$station_name, stations_messy$station_name, method = 'jw', p = 0.1)
row.names(dist.matrix.name) <- stations_messy$station_name
names(dist.matrix.name) <- stations_messy$station_name
dist.matrix.name <- as.dist(dist.matrix.name)
clusters.name <- hclust(dist.matrix.name, method = "ward.D2")
plot(clusters.name) # there is an error in the spelling of 'Main station'

# step 3b. fix the spelling error identified in step 3a
stations_clean_v2$station_name[stations_clean_v2$station_name == 'Main statoin'] <- 'Main station'

# step 3c. check that the spelling error was fixed
grep('Main statoin', stations_clean_v2$station_name) # the spelling error is no longer detected

# step 4. use a cluster diagram to look for typos in station_type
dist.matrix.type <- stringdist::stringdistmatrix(stations_messy$station_type, stations_messy$station_type, method = 'jw', p = 0.1)
row.names(dist.matrix.type) <- stations_messy$station_type
names(dist.matrix.type) <- stations_messy$station_type
dist.matrix.type <- as.dist(dist.matrix.type)
clusters.type <- hclust(dist.matrix.type, method = "ward.D2")
plot(clusters.type) # there are no spelling errors

# step 5. check if any of the zeros are uppercase O's
str_detect(stations_messy$station_code, '[O]') # no uppercase O's
str_detect(stations_messy$waterbody_code, '[O]') # no uppercase O's

# step 6. check that all waterbody and station codes start with the correct sequence
str_detect(stations_messy$waterbody_code, '\\bWB0') # They all start with the correct sequence
str_detect(stations_messy$station_code, '\\bST0') # They all start with the correct sequence

# step 7. check that the dimensions of the cleaned dataset match the messy dataset
stations_clean_v2 %>% 
  verify(nrow(.) == 97) # no error message means the number of rows is the same
stations_clean_v2 %>% 
  verify(ncol(.) == 6) # no error message means the number of columns is the same
```





### 3c. Waterbodies dataset

From my exploration of the waterbodies dataset above, I have identified the following errors:
  *Individual outliers*
    - no previously identified errors
  
  *Formatting errors*
    - no previously identified errors
    
  *String errors*  
    - no previously identified errors
    
  *Invalid values*
    - no previously identified errors 
  
Since there are only two 'character' columns in this dataset, I can only look for *string errors*

Before I begin, I will duplicate the dataset so I'm not editing the raw data
```{r}
waterbodies_clean <- waterbodies_messy
```


```{r}
# step 1. look for spaces in columns where there shouldn't be
str_detect(waterbodies_messy$waterbody_code, '[[:space:]]') # no spaces

# step 2. look for lowercase letters in columns where there shouldn't be 
str_detect(waterbodies_messy$waterbody_code, '[[:lower:]]') # no lowercases

# step 3a. look for lowercase letters at the beginning of names
str_view(waterbodies_messy$waterbody_name, '\\b[[:lower:]]') # Quinn Lake starts with a lowercase

# step 3b. fix the capitalization error found in step 3a
waterbodies_clean$waterbody_name[waterbodies_clean$waterbody_name == 'quinn lake'] <- 'Quinn Lake'

# step 3c. check that the capitalization error was fixed
grep('quinn lake', waterbodies_clean$waterbody_name) # the capitalization error is no longer present

# step 4. check if any of the zeros are uppercase O's
str_detect(waterbodies_messy$waterbody_code, '[O]') # no uppercase O's

# step 5. check that all waterbody codes start with "WB0"
str_detect(waterbodies_messy$waterbody_code, '\\bWB0') # They all start with the correct sequence

# step 6. check that the variable classes in the cleaned dataset are still correct
is.character(waterbodies_clean$waterbody_code)
is.character(waterbodies_clean$waterbody_name)

# step 7. check that the dimensions of the clean dataset match the messy dataset
waterbodies_clean %>% 
  verify(nrow(.) == 87) # no error message means the number of rows is the same
waterbodies_clean %>% 
  verify(ncol(.) == 2) # no error message means the number of columns is the same
```




### 3d. Match Records Across Tables*

*Identify Shared Columns*
  - waterbody_code is the shared column between the stations and waterbodies datasets
  - station_code is the shared column between the stations and chl a datasets

*Check Records*
Before I can join the tables, I have to check that records in one table have corresponding records in the related table. To do this, I'm going to compare each rows in the shared columns.

First, I will look at the waterbody_code columns. Since there are multiple stations per waterbody, I will have to filter out repeat rows in stations_clean_v2$waterbody_code
```{r}
# step 1. extract waterbody_code columns
stations_wb_code <- stations_clean_v2 %>%
  select(waterbody_code) %>%
  arrange(waterbody_code) %>%
  group_by(waterbody_code) %>%
  distinct()
waterbodies_wb_code <- waterbodies_clean %>%
  select(waterbody_code) %>%
  arrange(waterbody_code)

# step 2. compare to see if there are matching values
stations_wb_code$waterbody_code == waterbodies_wb_code$waterbody_code
```
All of the waterbody codes appear in both datasets


Next, I will look at the station_code columns. Since there are multiple samples per station, I will have to filter out repeat rows in chl_a_clean_v2$station_code
```{r}
# step 1. extract station_code columns
chl_a_st_code <- chl_a_clean_v2 %>%
  select(station_code) %>%
  arrange(station_code) %>%
  group_by(station_code) %>%
  distinct()
stations_st_code <- stations_clean_v2 %>%
  select(station_code) %>%
  arrange(station_code)

# step 2. compare to see if there are matching values
stations_st_code$station_code == chl_a_st_code$station_code # there are unmatched values because the datasets are two different lengths
```
There are a different number of station codes in each dataset, therefor not all of the stations in the stations_clean_v2 dataset occur in the chl_a_clean_v2 dataset.


*Investigate Unmatched Records*

I am first going to look at the number of unique station codes in in each dataset. Then, I will look at which stations are missing from the chl a dataset
```{r}
# step 1. determine the number of stations in each dataset
nrow(stations_st_code) # 97 rows = 97 stations
nrow(chl_a_st_code) # 70 rows = 70 stations

# step 2. create a new column in each dataset with just the station number
chl_a_st_code <- chl_a_st_code %>%
  mutate(station_ID = station_code) %>%
  separate(station_ID, into =  c("st", "ID"), sep = "T") %>%
  select(!st) %>%
  rename(station_code_chla = station_code)
stations_st_code <- stations_st_code %>%
  mutate(station_ID = station_code) %>%
  separate(station_ID, into =  c("st", "ID"), sep = "T") %>%
  select(!st) %>%
  rename(station_code_st = station_code)

# step 3. determine which stations are missing from the chl a dataset
stations_missing <- left_join(stations_st_code, chl_a_st_code, by = "ID") %>%
  filter(is.na(station_code_chla))
```
The difference in the number of rows from step 1 is 27, meaning 27 stations should be missing from chl_a_clean_v2. However, when I joined the two station_code columns together in steps 2 and 3, I get 28 stations that were missing from chl_a_clean_v2. Since all stations should be present in the stations dataset, it seems like I have an extra station present in the chl a dataset. 

To investigate this, I will do an anti-join to see which station from the chl a dataset is not present in the stations dataset
```{r}
# step 1. anti_join the datasets to see which one is extra
anti_join(chl_a_st_code, stations_st_code, by = "ID")
```
The extra station is the one with the typo I corrected in section 3a. Chl a dataset, so it seems like i corrected the typo wrong. There is no way to know what the correct station name should be since, the only shared column between the stations and chl a datasets in station_code. Since there is data in this row, I am not going to remove the entire row. I will leave this station code for now, so I can include this data in the joined table, but I will change the station_code to NA once my table is complete.



*Join the Tables*

I want my completed table to be based off of the information in the chl a dataset. 

First I need to joint the waterbodies and stations datasets together using the waterbody_code column. I want have one row for each station, so I need to join the waterbodies dataset into the stations dataset. Since the stations dataset has 97 rows, my joined table should also have 97 rows. The stations dataset has 6 columns and the waterbodies dataset has 2, therefor my joined table should have 7.
```{r}
# step 1. join the tables by waterbody_code
waterbodies_stations <- left_join(stations_clean_v2, waterbodies_clean, by = "waterbody_code")

# step 2. check that the table has the expected number of rows and columns
nrow(waterbodies_stations) # 97 rows as expected
ncol(waterbodies_stations) # 7 columns as expected
```

Then, I can join the waterbodies_stations dataset with the chl a dataset using the station_code column. I want to have one row for each chl a sample, so I need to join the waterbodies_stations dataset into the chl a dataset. Since the cleaned chl a dataset has 1577 rows, my joined table should also have 1577 rows. The chl a dataset has 15 columns and the waterbodies_stations dataset has 7, therefor my joined table should have 21. Once the tables are fully joined, I will also replace the incorrect station ID (ST099) identified above with NA
```{r}
# step 1. join the tables by station_code
chl_a_joined <- left_join(chl_a_clean_v2, waterbodies_stations, by = "station_code")

# step 2. check that the table has the expected number of rows and columns
nrow(chl_a_joined) # 1577 rows as expected
ncol(chl_a_joined) # 21 columns as expected

# step 3. replace incorrect station ID with NA
chl_a_joined$station_code[chl_a_joined$station_code == 'ST099'] <- NA

# step 4. check that the incorrect station ID was replaced
grep('ST099', chl_a_joined$station_code) # the incorrect ID is no longer present
```






